RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R

### 2014-07-25 manuscripts/manuscript.Rmd

In this paper, we introduce the reader to the concepts and use of
RNeXML and related technologies. We begin by discussing methods to read
phylogenetic tree data and character trait data in the NeXML format into
R, and also how to write phylogenetic data and character trait data into
NeXML. The heart of this section focuses on tools for reading and writing
the kind of semantically rich metadata that most sets NeXML apart from
other phylogenetic data formats. Starting simple, we use examples of
increasing complexity to illustrate concepts and tools from biodiversity
informatics that allow users to express and interpret complicated data
structures automatically.  We end with an example that uses metadata to
extend the NeXML standard to describe stochastic character mapping, one
of the kinds of data annotation that is currently available in R formats
but lacks a consistent, extensible export format. We illustrate how NeXML
files can be formally published to a data repository directly from R.
Finally, we provide an brief description of how the software has been
implemented, what quality control measures are in place, as well as
information about software reuse and getting support.

@@

@@

The resulting `nex` object is an R object representation of the NeXML. This is 
a native R object for representing phylogenetic trees (as well as 
character traits, and metadata), analogous to the `phylo` object from the 
popular `ape` R package `r citep(citation("ape"))`. While R functions and packages 
could be written to use this `nexml` object directly, most users will prefer to use
already existing methods provided by packages they know. From `nex` object we can 
therefore extract any phylogenies it contains in the `ape::phylo` format:

@@

We can then leverage the rich suite of methods available from the R
phylogenetics community. For instance, here we use the standard plotting function from
the `ape` package to plot the resulting `ape::phylo` object returned by
`get_trees`.

@@

The ability to read in phylogenetic trees in the NeXML format opens up a
wide and rapidly growing array of resources to the R user.  For instance,
a user can access all the phylogenies available in TreeBASE through the
`nexml` format.  `RNeXML` can read directly from a URL:

```{r eval=FALSE}
nex = nexml_read("https://raw.github.com/rvosa/supertreebase/master/data/treebase/S100.xml")
```

@@

```{r }
comp_analysis <- system.file("examples", "comp_analysis.xml", package="RNeXML")
nex <- nexml_read(comp_analysis)
get_characters(nex)
```

Returns a `data.frame` with columns as characters and rows as
taxa. Continuous and discrete characters are represented as separate
blocks in NeXML files, but will be combined as separate columns of a
single data frame by `get_characters()` if they correspond to the same
taxa. The `get_characters_list()` function will consistently receive
a list of data frames corresponding to each character matrix block of
the NeXML.

@@

To adjust these and other common metadata fields that should 
be part of any data deposition, `RNeXML` provides the `add_basic_meta()`
function.  This function does not assume any familiarity with the concepts
of namespaces or other more advanced metadata tools discussed later.  The
`add_basic_meta` takes an existing `nexml` object (or will create a new one
if none is provided, like the other `add_` functions), and any of the following
fields: `title`, `description`, `creator`, `pubdate`, `rights`, `publisher`,
`citation`.  Additional fields may be added to subsequent versions, see
`?add_basic_metadata` for the most up-to-date information.  Here we create
an new `nexml` object with the basic metadata for the `bird.orders` data
provided in the `ape` package:

@@

  description =" This data set describes the phylogenetic relationships of the
     orders of birds as reported by Sibley and Ahlquist (1990). Sibley
     and Ahlquist inferred this phylogeny from an extensive number of
     DNA/DNA hybridization experiments. The ``tapestry'' reported by
     these two authors (more than 1000 species out of the ca. 9000
     extant bird species) generated a lot of debates.

@@

or to papers by simply using the object's DOI (using `r citet(citation("knitcitations"))`):

@@

(Which is the paper on which the `geospiza` phylogeny from the geiger dataset is based)

@@

This function uses the `taxize` R library `r citep("Chamberlain taxize F1000Research 2013")` to check each taxon label against the NCBI database.
If a unique match is found, a metadata annotation is added to the taxon
providing the NCBI identification number to the taxonomic unit.  If no
match is found, the user is warned to check for possible typographic
errors in the taxonomic labels provided.

@@

So far our examples have relied on functions that have a built-in
understanding of the kind of metadata being provided, and can thus
generate the appropriate metadata with little further guidance.
In order to be machine-readable and understandable, NeXML requires
all metadata annotations follow the RDFa format (see @Vos_2012) which
explicitly declares the meaning of the annotation in a way a computer
can understand.  For instance, the title we gave to the `bird.orders`
phylogeny appears in the NeXML as:

@@

Of course writing out such a definition manually becomes tedious quickly. Because
these are just R commands, we can easily define a function that can loop over an
assignment like this for each edge, extracting the appropriate order, length and
state from an existing R object such as that provided in the `phytools` package.  
Likewise, it is straight forward to define a function that reads this data using
the `RNeXML` utilities and converts it back to the `phytools` package. The full
implementation of this mapping can be seen in the `simmap_to_nexml()` and the
`nexml_to_simmap()` functions provided in the `RNeXML` package.

>>>>

In this paper, we introduce the reader to the concepts and use of
RNeXML and related technologies. We begin by discussing methods to read
phylogenetic tree data and character trait data in the NeXML format into
R, and also how to write phylogenetic data and character trait data into
NeXML. The heart of this section focuses on tools for reading and writing
the kind of semantically rich metadata that most sets NeXML apart from
other phylogenetic data formats. Starting simple, we use examples of
increasing complexity to illustrate concepts and tools from biodiversity
informatics that allow users to express and interpret complicated data
structures automatically.  We end with an example that uses metadata to
extend the NeXML standard to describe stochastic character mapping, one
of the kinds of data annotation that is currently available in R formats
but lacks a consistent, extensible export format. We illustrate how NeXML
files can be formally published to a data repository directly from R.
Finally, we provide a brief description of how the software has been
implemented, what quality control measures are in place, as well as
information about software reuse and getting support.

@@

<!-- I think it would be useful to provide an overview of what is contained in this -->
<!-- example file here. The reader will know better what to expect (e.g., the -->
<!-- object contains 2 trees.) -->

@@

The resulting `nex` object is an R object representation of the NeXML. This is 
a native R object, of the `nexml` class, for representing phylogenetic trees (as well as 
character traits, and metadata), analogous to the `phylo` object from the 
popular `ape` R package `r citep(citation("ape"))`. While R functions and packages 
could be written to use this `nexml` object directly, most users will prefer to use
already existing methods provided by packages they know. From a `nexml` object, we can 
therefore extract any phylogenies it contains in the `ape::phylo` format:

@@

We can then leverage the rich suite of methods available from the R
phylogenetics community. For instance, here we use the standard plotting function from
the `ape` package to plot the first tree in the resulting `ape::phylo` object returned by
`get_trees`.

<!-- I am not sure it really matters but `phy` is actually a `multiPhylo` -->
<!-- object -->

@@

The ability to read in phylogenetic trees in the NeXML format opens up a
wide and rapidly growing array of resources to the R user.  For instance,
a user can access all the phylogenies available in TreeBASE <!-- first mention -->
<!-- of TreeBase, include URL? or introduce it in intro? --> through the
`nexml` format.  `RNeXML` can read directly from a URL:

```{r eval=FALSE}
nex <- nexml_read("https://raw.github.com/rvosa/supertreebase/master/data/treebase/S100.xml")
```

@@

```{r }
comp_analysis <- system.file("examples", "comp_analysis.xml", package="RNeXML")
nex <- nexml_read(comp_analysis)
get_characters(nex)
```
<!-- I think it's better to be explicit rather than referring to the chunk above -->

`get_characters` returns a `data.frame` with columns as characters and rows as
taxa. Continuous and discrete characters are represented as separate
blocks in NeXML files, but will be combined as separate columns of a
single data frame by `get_characters()` if they correspond to the same
taxa. The `get_characters_list()` function will consistently receive
a list of data frames corresponding to each character matrix block of
the NeXML.

@@

To adjust these and other common metadata fields that should 
be part of any data deposition, `RNeXML` provides the `add_basic_meta()`
function.  This function does not assume any familiarity with the concepts
of namespaces or other more advanced metadata tools discussed later.  The
`add_basic_meta` takes an existing `nexml` object (or will create a new one
if none is provided, like the other `add_` functions), and any of the following
fields: `title`, `description`, `creator`, `pubdate`, `rights`, `publisher`,
`citation`.  Additional fields may be added to subsequent versions, see
`?add_basic_metadata` for the most up-to-date information.  Here we create
a new `nexml` object with the basic metadata for the `bird.orders` data
provided in the `ape` package:

@@

  description = "This data set describes the phylogenetic relationships of the
     orders of birds as reported by Sibley and Ahlquist (1990). Sibley
     and Ahlquist inferred this phylogeny from an extensive number of
     DNA/DNA hybridization experiments. The ``tapestry'' reported by
     these two authors (more than 1000 species out of the ca. 9000
     extant bird species) generated a lot of debates.

@@

or to papers by simply using the object's DOI (using `r citet(citation("knitcitations"))`).
For instance to add the citation information of the paper that generated the
`geospiza` phylogeny included in the `geiger` package:

@@

@@

This function uses the R package `taxize` `r citep("Chamberlain taxize F1000Research 2013")` to check each taxon label against the NCBI database.
If a unique match is found, a metadata annotation is added to the taxon
providing the NCBI identification number to the taxonomic unit.  If no
match is found, the user is warned to check for possible typographic
errors in the taxonomic labels provided.

@@

So far our examples have relied on functions that have a built-in
understanding of the kind of metadata being provided, and can thus
generate the appropriate metadata with little further guidance.
In order to be machine-readable and understandable, NeXML requires
all metadata annotations follow the RDFa format (see @Vos_2012) which
explicitly declares the meaning of the annotation <!-- in a way a computer
can understand -- this is redundant with the beginning of the sentence -->.  For instance, the title we gave to the `bird.orders`
phylogeny appears in the NeXML file as:

@@

Of course writing out such a definition manually becomes tedious quickly. Because
these are just R commands, we can easily define a function that can loop over an
assignment like this for each edge, extracting the appropriate order, length and
state from an existing R object such as that provided in the `phytools` package.  
Likewise, it is straightforward to define a function that reads this data using
the `RNeXML` utilities and converts it back to the `phytools` package. The full
implementation of this mapping can be seen in the `simmap_to_nexml()` and the
`nexml_to_simmap()` functions provided in the `RNeXML` package.
